Epigram 2\\ more design thoughts

### 2010-08-31 notes/Epigram2/Epigram2.tex
notes: fixed reversal, finished description of example Those of you who guessed that I had a lapse of concentration when I wrote that you could extract a .gram from an executable can pat yourselves on the back. I meant the other way around and have repaired the damage. I've also added some superficial description of the example program. Tighter definitions are yet to come. The purpose of this section is to give an impression of the interaction.

A separate (or at least separable) component, based on Edwin Brady's
\textsc{epic} compiler, can generate a \texttt{.gram} file from an
executable. The latter, if executed, offers a straightforward
read-eval-print loop for first-order expressions constructed over the
signature of definitions provided by the original source. Incomplete
definitions in the source may result in run-time errors: these are the
only run-time errors.  Such an executable, invoked with an argument,
treats that argument as a single command session.

@@

A \textbf{document} is a series of \textbf{developments}. A
development comprises a \textbf{declaration} introducing a problem,
and a \textbf{refinement} reducing the problem to a hopefully
empty collection of subproblems. Too much abstraction: example, please!
\newcommand{\capbox}[2]{\raisebox{0.12in}{\begin{array}[t]{|l|}\hline
      #1 \vspace*{-0.05in}\\
    \hfill\mbox{\scriptsize{#2}} \\ \hline\end{array}}}
\newcommand{\Ts}[1]{\texttt{#1}\;}
\newcommand{\co}{\;:\;}
\newcommand{\sco}{;\;\;}
\newcommand{\cm}{,\;}
\[
\capbox{
  \Ts{let}
    \capbox{
      \capbox{\Ts{x} \cm \Ts{y}\co \Ts{Nat}}{declaration}\\
      \Ts{--------------}\\
      \capbox{\Ts{x} \Ts{+} \Ts{y}}{template} \co \capbox{\Ts{Nat}}{type}
      }
      {declaration}\\ \\
  \capbox{
     \capbox{\Ts{x}\Ts{+}\Ts{y}}{problem}
       \capbox{\Ts{<=}\capbox{\Ts{induction}\Ts{x}}{eliminator}}{tactic}\\
     \capbox{
      \{\;\capbox{\capbox{\Ts{'zero}\Ts{+}\Ts{y}}{problem}
                  \capbox{\Ts{=} \capbox{\Ts{y}}{term}}{tactic}
                  \capbox{}{block}
          }{refinement}\\
      ;\;\capbox{\capbox{\Ts{'suc}\Ts{x}\Ts{+}\Ts{y}}{problem}
                 \capbox{\Ts{=} \capbox{\Ts{'suc}
                   \capbox{\{!\;!\}}{shed}}{term}}{tactic}
                  \capbox{}{block}
         }{refinement}\\
      \}
      }{block}
    }
    {refinement}}
    {development}
\]

>>>>

A separate (or at least separable) component, based on Edwin Brady's
\textsc{epic} compiler, can generate an executable from a \texttt{.gram}
file. The former, if executed, offers a straightforward
read-eval-print loop for first-order expressions constructed over the
signature of definitions provided by the original source. Incomplete
definitions in the source may result in run-time errors: these are the
only run-time errors.  Such an executable, invoked with an argument,
treats that argument as a single command session.

@@

A \textbf{document} is a series of \textbf{developments}. A
development comprises a \textbf{declaration} introducing a problem,
and a \textbf{refinement} reducing the problem to a hopefully
empty collection of subproblems. Too much abstraction: example, please!
\newcommand{\capbox}[2]{\raisebox{0.12in}{\begin{array}[t]{|l|}\hline
      #1 \vspace*{-0.05in}\\
    \hfill\mbox{\scriptsize{#2}} \\ \hline\end{array}}}
\newcommand{\Ts}[1]{\texttt{#1}\;}
\newcommand{\co}{\;:\;}
\newcommand{\sco}{;\;\;}
\newcommand{\cm}{,\;}
\[
\capbox{
  \Ts{let}
    \capbox{
      \capbox{\Ts{x} \cm \Ts{y}\co \Ts{Nat}}{declaration}\\
      \Ts{--------------}\\
      \capbox{\Ts{x} \Ts{+} \Ts{y}}{template} \co \capbox{\Ts{Nat}}{type}
      }
      {declaration}\\ \\
  \capbox{
     \capbox{\Ts{x}\Ts{+}\Ts{y}}{problem}
       \capbox{\Ts{<=}\capbox{\Ts{induction}\Ts{x}}{eliminator}}{tactic}\\
     \capbox{
      \{\;\capbox{\capbox{\Ts{'zero}\Ts{+}\Ts{y}}{problem}
                  \capbox{\Ts{=} \capbox{\Ts{y}}{term}}{tactic}
                  \capbox{}{block}
          }{refinement}\\
      ;\;\capbox{\capbox{\Ts{'suc}\Ts{x'}\Ts{+}\Ts{y}}{problem}
                 \capbox{\Ts{=} \capbox{\Ts{'suc}
                   \capbox{\{!\;!\}}{shed}}{term}}{tactic}
                  \capbox{}{block}
         }{refinement}\\
      \}
      }{block}
    }
    {refinement}}
    {development}
\]

Lots of points show up from this example. I'll skim through them,
starting with the declaration: we give a template for addition,
hypothesizing declarations for the fields of the template, then giving
a type for the whole thing. This template determines the initial
problem. You can tell which tokens in the template are punctuation
rather than placeholders, because they don't have declarations!
Note that here, we can see value declarations, but we shall also have
propositional declarations, \(\texttt{:-}\;P\), where a proposition is
asserted anonymously---\(P\) then gives the shapes of problems for
refinement, i.e., propositional goals.

Refinements go `\textbf{problem} \textbf{tactic} \textbf{block}'. A
refinement is an offer to solve a problem of a particular form.
Correspondingly, the elaborator needs to check that the refinement
offered suits the problem or problems at hand: the rigidity of this
check will soften with time, but the minimum acceptable level is that
the refinement's problem freely determines the programmer's choice of
names for the variables in the problem. Working interactively, of course,
the problem parts of refinements are generated mechanically from the
problems at hand, ensuring a match.

A tactic provides the means to split a problem into zero or more
subproblems. The \texttt{<=} (`by') tactic invokes an \textbf{eliminator}
in the elimination-with-a-motive style. The \texttt{=} (`return') tactic
takes a term explaining how to compute the return value, and leaves no
subproblems. There will doubtless be others, not least the notorious
\texttt{with}.

The \textbf{shed}, \texttt{\{! !\}}, stands in place of a piece of program
(most syntactic categories, not just terms) to indicate that the programmer
has yet to choose something, thus causing elaboration to suspend. Note
that a tactic \texttt{= \{! !\}} does commit to return, where a
shed directly in a tactic place does not.
